The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2021-03-23 Manuscript/NBABubbleAnalysis.Rmd
Update NBABubbleAnalysis.Rmd

>>>>

# Appendix {#sec:Appendix}
```{r, echo=FALSE}
permutation_test <- function(data1, data2, test_column_num, n, R) {
  #calculate test statsitic and store it
  z <- z.test(x=data1[ ,test_column_num], y=data2[ ,test_column_num], sigma.x =
                sd(data1[ ,test_column_num]), sigma.y = sd(data2[ ,test_column_num]))
  z <- abs(z$statistic)

  #calculate mean of combined samples
  overallmean<-c(data1[ ,test_column_num], data2[ ,test_column_num])
  overallmean<-mean(overallmean)

  #transform datasets using xi*=xi-x̄-z̄
  data1transformed <- data1[ ,test_column_num] - mean(data1[ ,test_column_num]) + overallmean
  data1transformed <- data.frame(data1transformed)
  data2transformed <- data2[ ,test_column_num] - mean(data2[ ,test_column_num]) + overallmean
  data2transformed <- data.frame(data2transformed)

  #collect R samples of n data points from each group and calculate test statistic between each pair of samples
  bootstrap <- data.frame("Bootsrap T-stat"=numeric(), "Condition"=numeric())
  set.seed(1234)
  for (i in 1:R){
    sample1 <- data1transformed[sample(1:nrow(data1transformed), nrow(data1transformed), replace=TRUE),]
    sample2 <- data2transformed[sample(1:nrow(data2transformed), nrow(data2transformed), replace=TRUE),]

    bt <- z.test(x=sample1, y=sample2, sigma.x =
                  sd(sample1), sigma.y = sd(sample2))
    bts <- abs(bt$statistic)

    bootstrap[i, ] <- bts
    #determine if bootstrap statistics is larger than observed and save result
    if(bootstrap[i,1] >= z) {bootstrap[i,2]=1} else {bootstrap[i,2]=0}
  }
  #calculate proportion of bootsrap t-statistics wihch are larger than or equal to observed test statistic
  p_value <- sum(bootstrap[, 2])/nrow(bootstrap)
  return(p_value)
}

row_one <- permutation_test(game_log_2020_home, game_log_2017_2019_home, 6, 100, 10000)
row_two <-permutation_test(game_log_2020_home, game_log_2017_2019_home, 8, 100, 10000)
row_three <- permutation_test(game_log_2020_away, game_log_2017_2019_away, 8, 100, 10000)
row_four <- permutation_test(game_log_2020_home, game_log_2017_2019_home, 14, 100, 10000)
row_five <- permutation_test(game_log_2020_home, game_log_2017_2019_home, 17, 100, 10000)
row_six <- permutation_test(game_log_2020_home, game_log_2017_2019_home, 20, 100, 10000)
row_seven <-  permutation_test(game_log_2020_away, game_log_2017_2019_away, 14, 100, 10000)
row_eight <- permutation_test(game_log_2020_away, game_log_2017_2019_away, 17, 100, 10000)
row_nine <- permutation_test(game_log_2020_away, game_log_2017_2019_away, 20, 100, 10000)
perm_test_res <- c(row_one, row_two, row_three, row_four, row_five, row_six, row_seven, row_eight, row_nine)
perm_test_col <- matrix(data=perm_test_res, ncol = 1)
app_resultsfinal <-cbind(resultsfinal, perm_test_col)
colnames(app_resultsfinal)[5] <- "Bootstrap Z-test"
```

In an effort to advance and strengthen the paper we conducted bootstrap analysis for every
hypothesis. A custom R function was developed and utilized to conduct this analysis, the code for 
that function can be found below:

```{r, echo = TRUE, eval=FALSE}
permutation_test <- function(data1, data2, test_column_num, n, R) {
  #calculate test statsitic and store it
  z <- z.test(x=data1[ ,test_column_num], y=data2[ ,test_column_num], sigma.x =
                sd(data1[ ,test_column_num]), sigma.y = sd(data2[ ,test_column_num]))
  z <- abs(z$statistic)

  #calculate mean of combined samples
  overallmean<-c(data1[ ,test_column_num], data2[ ,test_column_num])
  overallmean<-mean(overallmean)

  #transform datasets using xi*=xi-x̄-z̄
  data1transformed <- data1[ ,test_column_num] - mean(data1[ ,test_column_num]) + overallmean
  data1transformed <- data.frame(data1transformed)
  data2transformed <- data2[ ,test_column_num] - mean(data2[ ,test_column_num]) + overallmean
  data2transformed <- data.frame(data2transformed)

  #collect R samples of n data points from each group and calculate test statistic between each pair of samples
  bootstrap <- data.frame("Bootsrap T-stat"=numeric(), "Condition"=numeric())
  set.seed(1234)
  for (i in 1:R){
    sample1 <- data1transformed[sample(1:nrow(data1transformed), nrow(data1transformed), replace=TRUE),]
    sample2 <- data2transformed[sample(1:nrow(data2transformed), nrow(data2transformed), replace=TRUE),]

    bt <- z.test(x=sample1, y=sample2, sigma.x =
                  sd(sample1), sigma.y = sd(sample2))
    bts <- abs(bt$statistic)

    bootstrap[i, ] <- bts
    #determine if bootstrap statistics is larger than observed and save result
    if(bootstrap[i,1] >= z) {bootstrap[i,2]=1} else {bootstrap[i,2]=0}
  }
  #calculate proportion of bootsrap t-statistics wihch are larger than or equal to observed test statistic
  p_value <- sum(bootstrap[, 2])/nrow(bootstrap)
  return(p_value)
}
```

The results from the bootstrap tests can be compared to the previous tests results 
in order to strengthen understanding. These comparisons can be made using the table, Table 
\@ref(tab:table2), below.

```{r table2, echo = FALSE}
knitr::kable(app_resultsfinal, caption = 'Previous Results Side By Side With Bootstrap Results', align = "cccccc", booktabs = TRUE)
```

As you can see in the table, the p-values for the tests using bootstrapping
align closely with p-values from the other tests. Starting with the home team
win percentage, we see our p-value is 0.0405. This is important because 
earlier in our paper we found that the p-values for z-test(.0497) and Fisher’s
exact test(.0400) were initially significant, but not after a Bonferroni 
correction. Thus, the bootstrapping result gives some extra level of confidence
to the conclusion that the home team winning percentage was different from
years past.

For home teams, the bootstrapping tests seem to confirm the conclusion
playing in the NBA bubble had very little impact on performance. The bootstrap
tests for home scoring, two-point percentage and three-point percentage all
have very high p-values and agree with previous tests that there is no evidence
of a difference in these measures. One interesting finding is for the home team
free throw percentage. Earlier, we found the z-test (p=0.0692) was insignificant,
but the Wilcoxon test(p=0.0496) was significant. The bootstrapping test(p=0.551)
agrees with the z-test that there is no significant change in home team free throw
percentage. Since two of three tests run are insignificant and the third is very
close, it is probably safest to conservatively say that there is not much evidence
to support a claim that home team free throw percentage changed.

For away teams, we saw the bootstrapping test confirm the results of the previous
tests in every case. The bootstrapping methods corroborate the insignificant 
results for away three-point (p=0.3494) and away free throw shooting (p=0.4743).
Likewise, away team scoring(p=.0012) and two-point percentage(p=.0001) still had
highly significant p-values for the bootstrap test, just like the other tests for
these two measures. This gives us great confidence that away scoring and two-point
percentage changed in the bubble. If you recall, the confidence intervals for 
these measures support the claim that they increased. This is important because
it suggests when home-court advantage is removed, away teams were able to score 
more points by increased two-point efficiency, suggesting home-court advantage 
stems from an adverse effect on away teams. Recall that @Harris found that 
home-court advantage was maximized when home teams capitalized on their
advantage shooting the two-pointer. When away teams no longer face more 
challenging conditions for two-point shooting while going on the road, home
teams lose their advantage.

Overall, the addition of the bootstrapping methods to the study
appear to strengthen the conclusions made in the paper.
